---
title: "Models"
subtitle: "Project: Dominant Party Regimes in Sub-Saharan Africa"
format: 
  pdf:
    fig-pos: "H"
    fontsize: 11pt
    highlight-style: breezedark
    fig-width: 7
    fig-height: 5
    include-in-header: 
      - file: ../preamble.tex
execute: 
  echo: true
---

This file reports the modelling process and the associated analytic decisions. It also includes the results of sensitivity tests.

```{r}
#| label: setup
#| include: false

library(here)
source(here("utilities", "check_packages.R"))
source(here("utilities", "functions.R"))

theme_set(theme_minimal())
```

Read in the data compiled in the *data_building.qmd* file.

```{r}
#| label: read in data

data <- read_rds(here("data", "data_built", "data.built.rds"))
```

The main and only model in my original work was a multinomial logistic model. The main disadvantage of using this class of models in this context is that logistic models typically require more observations than linear models. Nevertheless, I start with a multinomial model to i) establish a baseline, ii) provide ground to validate the analytical coding of the dominant party regimes and iii) because there are not many alternative methods to analyze the data.

I first use a preferred specification, with party system institutionalization and education variables imputed with the mean of 2013-2018 values for the country with missing data, ELF imputed with previous values for the country and oil rents variable presented as a dummy.

```{r}
#| label: multinomial model, prefered specification

m1a <- multinom(party_reg ~ v2xps_party_mean + elf_fill + col.type
               + housesys + v2x_polyarchy + oil_rent_dummy + e_peaveduc_mean,
               data = data)

m1b <- multinom(relevel(party_reg, ref = "Autocratic Dominant") ~ v2xps_party_mean + elf_fill + col.type
               + housesys + v2x_polyarchy + oil_rent_dummy + e_peaveduc_mean,
               data = data)

modelsummary(list(m1a, m1b), stars = T, shape = term ~ response, coef_rename = T)
```

